---
title: "Trabalho de Séries Temporais"
output: html_notebook
---

```{r}

# Import libraries that will be used in this project
#install.packages("portes")
library(tseries)
library(forecast)
library(fma)
library(ggplot2)
library(FitAR)
require(astsa)
library(nortest)
library(portes)
library(lattice)
library(tsutils)
library(TSA)
```

```{r}
# Import dataset
library(readxl)
dataset <- read_excel("/Users/paulaeveling/Desktop/TS-Project/dataset.xlsx")
```

```{r}
# Format dataset as time series
ds <- ts(dataset, start=c(1999,1), frequency=12)
```

```{r}
ds
```


```{r}
# Statistical summary of the dataset
summary(ds)
```

Based on the statistical summary of the dataset, it can be observed that the average unemployment rate between 1999 and 2021 is 9.74%, while the minimum observed was 4.9% and the maximum 18.65%.

```{r}
# Time Series Initial Analysis
tsdisplay(ds, main = "Unemployment Rate in Portugal from 1999 until 2021")

plot(ds, main = "Unemployment Rate in Portugal from 1999 until 2021")
```
Trend: The dataset presents an upward trend until 2013, when the unemployment rate drops.
Seasonality: It is not clear in the charts if this is a seasonal time series.
Cycle: There is no clear cicles in the dataset.
Variance: The variance is not constant.
Stationarity: The dataset does not present characteristics of a stationary time series.

```{r}
# Time series decomposition
decompose(ds)
plot(decompose(ds))
```

```{r}
# Seasonal Analysis
seasplot(ds, outplot=2)
```
Statistical Test:
H0: no seasonality
H1: there is seasonality
Since the p-value is lower than 0.05, there is enough evidence to say that this time series has seasonality.

It can be observed on the chart that the dataset under study is seasonal because there are boxes below the line, confirming the statistical test presented on the last step.

```{r}
# Seasonal Analysis
seasplot(ds, outplot=3)
```
Statistical Test:
H0: no seasonality
H1: there is seasonality
Since the p-value is lower than 0.05, there is enough evidence to say that this time series has seasonality.

It can be observed on the chart that the dataset under study is seasonal because there are boxes below the line, confirming the statistical test presented on the last step.

```{r}
# Stationarity Ttest

# Dickey-Fuller
adf.test(ds)

# KPSS
kpss.test(ds)

#
PP.test(ds) 
```

The Dickey-Fuller test presented a p-value higher than 0.1, so we fail to reject the null hypothesis. This is a non-stationary time series.

```{r}
# Stationarity test with one order differencing
adf.test(diff(ds))

# plot the timeseries with one differencing order
tsdisplay(diff(ds), main = "Unemployment Rate in Portugal from 1999 until 2021")
```
After applying one order differencing, the p-value presented by the Dickey-Fuller test is lower than 0.01, so we do reject the null hypothesis. There is strong evidence that the time-series is stationary.

It can be observed a strong seasonality in the time-series that must be treated accordingly.

```{r}
# Variance treatment using lambda
BoxCox(ds, interval=c(-4,4))
lambda.est<-0.449
```

```{r}
# Tratamento da Time Series Using 1 stationary differencing + 1 seasonal differencing + BoxCox
ds_treated<-ts(diff(diff(ds^lambda.est, lag=12, differences = 1)), frequency=12)
tsdisplay(ds_treated, lag=12, differences=1)
```
```{r}
# Auto ARIMA Fit
fit.auto<-auto.arima(ds, stationary=FALSE, lambda=lambda.est, seasonal = TRUE)
summary(fit.auto)
```

```{r}
fit1<- Arima(ds, order=c(1,2,0), seas=list(order=c(2,0,0), period=12),
             include.drift=TRUE, lambda=lambda.est)
summary(fit1)
```


```{r}
fit2<- Arima(ds, order=c(1,1,0), seas=list(order=c(2,1,0), period=12),
             include.drift=TRUE, lambda=lambda.est)
summary(fit2)
```

```{r}
fit3<- Arima(ds, order=c(1,2,1), seas=list(order=c(1,1,1), period=12),
             include.drift=TRUE, lambda=lambda.est)
summary(fit3)
```



```{r}
# Residual Analysis
tsdisplay(fit.auto$residuals)
tsdisplay(fit1$residuals)
tsdisplay(fit2$residuals)
tsdisplay(fit3$residuals)
```

```{r}
# Test for normality
lillie.test(fit2$residuals) #res?duo branco Gaussiano
```

```{r}
t.test(fit2$residuals)
```


```{r}
LjungBox(fit2$residuals, lags=seq(4,30,2), order=3, season=1, squared.residuals=FALSE)
```

```{r}
LiMcLeod(fit2$residuals, lags=seq(4,30,2), order=3, season=1, squared.residuals=FALSE)
```

```{r}
BoxPierce(fit2$residuals,lags=seq(4,30,2),season=1)
```

```{r}
# Plot adjusted model
plot(ds)
lines(fitted(fit2), col="red")
```


```{r}
# 24-month forecasting
forecast(fit2,h=24) 
plot(forecast(fit2,h=24))
lines(fitted(fit2),col=2)
```

CROSSCORRELATION (COVARIANCE) FUNCTION

```{r}
# importing dataset
#perspectiva <- read_excel("/Users/paulaeveling/Desktop/TS-Project/perspetiva.xlsx")

# importing dataset
perspectiva <- read_excel("/Users/paulaeveling/Desktop/TS-Project/two_datasets.xlsx")
```

```{r}
# transforming dataset into ts format
datasets <- ts(perspectiva, start=c(1999,1), frequency=12)
unemployment <- ts(perspectiva$unemployment, start=c(1999,1), frequency=12)
perspective <- ts(perspectiva$economic_perspective, start=c(1999,1), frequency=12)

# plot two time series
xyplot(datasets)
```

```{r}
cor(datasets)
```
```{r}
plot(decompose(perspective))
```

```{r}
length(unemployment)
length(perspective)
```
Both datasets have the same lenght, which is good.
Bots time series must be treated accordingly. The unemployment one needs one box cox transformation to treat the variance, one differencing to de-trend the series and one seasonal differencing to handle seasonality.

The economic perspective time series on the other hand, only needs one differencing to transform the series into stationary.

```{r}
ccf(x=perspective,y=unemployment)
```
```{r}
tsdisplay(perspective)
```
It approaches zero very slowly.

```{r}
treatedunemployment<-ts(diff(diff(unemployment^lambda.est, lag=12, differences = 1)), frequency=12)
treatedperspectiva<-ts(diff(perspective),frequency=12)
```

```{r}
length(treatedunemployment)
length(treatedperspectiva)
```

Because they both needed different treatments, we have lost 12 observations in the unemployment dataset and only one in the economic perspective.

```{r}
treatedunemployment
treatedperspectiva
```

Since both time series have different dizes, it is necessary to remove observations from the economic perspective so they both have the same size.

```{r}
treatedperspectiva2<-ts((treatedperspectiva[c(-1,-2,-3,-4,-5,-6,-7,-8,-9,-10,-11,-12)]), frequency=12)

treatedunemployment
treatedperspectiva2
```

```{r}
print(ccf(x=treatedperspectiva2,y=treatedunemployment))
abline(v=0, lty = 2)
```
We want to explain the variable unemployment using the variable economic perspective. No clear leading or lagging pattern! And also it is clear that both series have trend. Both are non-stationary.

```{r}
lag2.plot(treatedperspectiva2, treatedunemployment, 10)
```


```{r}
perspective_cbind3<-cbind(xlag3=lag(perspective,-1))
fit.regauto1<- auto.arima(unemployment, xreg=perspective_cbind3) # modelo automático.
fit.regauto1
```
```{r}
perspective_cbind2<-cbind(xlag3=lag(perspective,-2))
fit.regauto1<- auto.arima(unemployment, xreg=perspective_cbind2,d=1) # modelo automático.
fit.regauto1
```

```{r}
length(xreg_perspet)
length(unemployment)
```
```{r}
perspective_cbind2
```
```{r}
perspective_cbind3
```


```{r}
fit.regauto1<- auto.arima(unemployment, xreg=xreg_perspet) # modelo automático.
fit.regauto1
```



```{r}
treatedperspectiva23<-cbind(xlag6=lag(treatedperspectiva2,-3))
auto.arima(treatedunemployment,xreg=treatedperspectiva23,d=0)
```

Working with the differenced time series
```{r}

tsdisplay(auto.arima(treatedunemployment,xreg=treatedperspectiva23))
```


```{r}
tsdisplay(auto.arima(treatedunemployment,xreg=treatedperspectiva23)$residuals)
```

```{r}
tsdisplay(auto.arima(treatedunemployment,xreg=treatedperspectiva23))
```
diffperspective-t=0.7533∗diffunemployment+2+nt

```{r}
arimax(treatedunemployment,order=c(1,0,1),xreg=treatedperspectiva23)
```

PERIODOGRAM

```{r}
# importing dataset
#perspectiva <- read_excel("/Users/paulaeveling/Desktop/TS-Project/perspetiva.xlsx")

# importing dataset
perspectiva <- read_excel("/Users/paulaeveling/Desktop/TS-Project/two_datasets.xlsx")
```

```{r}
# transforming dataset into ts format
datasets <- ts(perspectiva, start=c(1999,1), frequency=12)
unemployment <- ts(perspectiva$unemployment, start=c(1999,1), frequency=12)
perspective <- ts(perspectiva$economic_perspective, start=c(1999,1), frequency=12)

treatedunemployment<-ts(diff(diff(unemployment^lambda.est, lag=12, differences = 1)), frequency=12)
treatedperspectiva<-ts(diff(perspective),frequency=12)
```

```{r}
require(TSA)
y <- treatedunemployment
length(treatedunemployment)
new = periodogram(y)
```


```{r}
order(new$spec) # approximate the Fourir frequencies
```

```{r}
new$freq[36]
(1/new$freq[36])
(1/new$freq[36])/12 # the harmonic frequencies
```


```{r}
treatedperspectiva2<-ts((treatedperspectiva[c(-1,-2,-3,-4,-5,-6,-7,-8,-9,-10,-11,-12)]), frequency=12)

x<-treatedperspectiva2
length(treatedperspectiva)
new = periodogram(x)
```

```{r}
order(new$spec) # approximate the Fourir frequencies
```

```{r}
new$freq[67]
(1/new$freq[67])
(1/new$freq[67])/12
```

